### 2017-07-09 related.tex
comments

\makeatletter{}\RequirePackage{fix-cm}
\documentclass[smallextended]{svjour3}

\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{comment}
\usepackage{mathrsfs}
\usepackage{balance}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[noend]{algpseudocode}
\usepackage[group-separator={,}]{siunitx}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{bm}
\usepackage[all]{hypcap}
\usepackage[absolute]{textpos}
\usepackage{wrapfig}
\usepackage[title]{appendix}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\LineComment}[1]{\State{\(\triangleright\) #1}}
\newcommand*\Let[2]{\State #1 := #2}

\smartqed

\usepackage{graphicx}
\graphicspath{{./figures/}}

\hypersetup{  bookmarks=true,
  unicode=true,
  pdftoolbar=true,
  pdfmenubar=true,
  pdffitwindow=true,
  pdfstartview={FitV},
  pdfnewwindow=true,
  colorlinks=false,
  pdfdisplaydoctitle=true,
  pdfborder={0 0 0}
}

\setlength{\TPHorizModule}{1in}
\setlength{\TPVertModule}{1in}
\textblockorigin{0.75in}{0.875in}

\setlist[itemize]{leftmargin=*,partopsep=5pt}
\setlist[enumerate]{leftmargin=*,partopsep=5pt}

\paperheight 11in
\paperwidth 8.5in

\usepackage{dcolumn}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\newcommand\nil{\textit{nil}}

\usepackage{tikz}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
      \node[shape=circle,draw,inner sep=0pt] (char) {#1};}}

\makeatletter
\newcommand*{\rome}[1]{\expandafter\@slowromancap\romannumeral #1@}
\makeatother

\makeatletter{}
\newcommand{\wifi}{Wi-Fi}
\newcommand{\codename}{\textsc{Verifier}}
\newcommand{\us}{$\mu$s}
\newcommand{\ns}{\textsc{NS-3}}

\newcommand{\shuvendu}[1]{{\bf [[SKL:{#1}]]}}

\def\thetitle{Wireless Protocol Validation Under Uncertainty}
\def\theauthors{}

\authorrunning{J. Shi, S. K. Lahiri, R. Chandra, G. Challen}

\journalname{Formal Methods in System Design}

\begin{document}

\title{  \thetitle\thanks{This work was published, in part, in Runtime Verification
    (RV) 2016~\cite{Shi2016}.}
}

\author{  Jinghao Shi$^1$ \and
  Shuvendu K. Lahiri$^2$ \and
  Ranveer Chandra$^2$ \and
  Geoffrey Challen$^1$
}

\institute{
  Jinghao Shi\\
  jinghaos@buffalo.edu\\\\
  Shuvendu K. Lahiri\\
  shuvendu@microsoft.com\\\\
  Ranveer Chandra\\
  ranveer@microsoft.com\\\\
  Geoffrey Challen\\
  challen@buffalo.edu\\\\
  $^1$University at Buffalo, Buffalo, NY 14260, USA\\
  $^2$Microsoft Research, Redmond, WA 98052, USA
}

\hypersetup{  pdfinfo={    Title={\thetitle},
    Author={\theauthors},
  }
}

\maketitle

\makeatletter{}\begin{abstract}
  Runtime validation of wireless protocol implementations cannot always employ
  direct instrumentation of the device under test (DUT). The DUT may not
  implement the required instrumentation, or the instrumentation may alter the
  DUT's behavior when enabled. Wireless sniffers can monitor the DUT's
  behavior without instrumentation, but they introduce new validation challenges.
  Losses caused by wireless propagation prevent sniffers
  from perfectly reconstructing the actual DUT packet trace. As a result,
  accurate validation requires distinguishing between
  specification deviations that represent implementation errors and those
  caused by sniffer uncertainty.

  \sloppy{
  We present a new approach enabling sniffer-based validation of wireless
  protocol implementations. Beginning with the original protocol monitor state
  machine, we automatically and completely encode sniffer uncertainty by
  selectively adding non-deterministic transitions. We characterize the
  NP-completeness of the resulting decision problem and provide an exhaustive
  algorithm for searching over all mutated traces. We also present
  practical protocol-oblivious heuristics for searching over the most likely
  mutated traces.  We have implemented our framework and show that it can
  accurately identify implementation errors in the face of uncertainty.
}

  \keywords{
    Runtime Verification \and
    Wireless Protocol \and
    Sniffer \and
  Uncertainty}
\end{abstract}

\makeatletter{}\section{Introduction}
\label{sec:intro}

Custom wireless protocols are often designed and deployed to meet the specific
performance and power needs of special-purpose wireless devices. Examples
include Google
Iris contact lenses~\cite{iris}, Xbox One wireless controllers~\cite{xbox}, and
Google Chromecast~\cite{chromecast}. Validating that device implementations work
correctly is critical to achieve the design goals of the wireless protocol and
also prevent bugs in shipped products~\cite{wifried,lollipop,surface}.

Runtime validation of the protocol implementations on such devices is
challenging because collecting traces from the device under test (DUT) is often
infeasible. The resource limitations of embedded or battery-powered devices may
cause them to not provide trace collecting capabilities. DUT may contain
proprietary hardware or firmware that hides the implementation details and
prevents testers from collecting traces through source code instrumentation.
Even when collecting trace directly from the DUT is possible, the overhead it causes
may alter the behavior of the DUT due to the observer
effect~\cite{mytkowicz2008observer}, threatening the validation results.

An attractive alternative is to use wireless
sniffers to record traffic generated by the DUT during testing.
Sniffers do not require direct access to the DUT or the need to alter its behavior.
However, due to the fundamentally unpredictable nature of wireless
communications, the packets captured by the sniffer will not exactly match
those received by the DUT.
The sniffer may miss packets that the DUT received, or receive packets that
the DUT missed.
This is true even when using multiple
sniffers~\cite{cheng2006jigsaw,mahajan2006analyzing,bahl2006enhancing}, a sniffer
with multiple antennas~\cite{omnipeek}, or in isolated wireless environments.

Since the sniffer trace may not perfectly match the actual trace, uncertainty
arises during protocol implementation validation.  For example, if the DUT fails
to respond correctly to a packet in the sniffer trace, it may either because the
DUT's implementation is incorrect, or the DUT did not actually receive the
packet, or the DUT's response was missed by the sniffer.  Whenever the DUT's
behavior does not match the specification, there are now two potential
explanations: either the DUT's implementation is wrong, or the sniffer trace is
inaccurate. Accurate validation requires  distinguishing between these two
causes.

We present a new technique than enables validation of protocol implementations using
wireless sniffers. Given a monitor state machine representing the protocol being
validated, we describe a systematic transformation that adds non-deterministic
transitions to incorporate uncertainty introduced by the sniffer. This
augmented validation state machine implicitly defines a set of
mutated traces, each satisfying the original state machine with a specific
likelihood. If the set is empty, the implementation definitely violates the
protocol. Searching over all the mutated traces is NP-complete, but the approach
can be made practical by applying protocol-oblivious heuristics that limit the
search to likely mutated traces.

Our paper makes the following contributions:
\begin{enumerate}
  \item To the best of our knowledge, we are the first to identify the
    uncertainty problem caused by sniffers in validating wireless protocol
    implementations.
  \item We formalize the problem using a nondeterministic state machine that
    systematically and completely encodes the uncertainty of the
    sniffer trace.
  \item We characterize the NP-completeness of the validation problem, and
    present two protocol-oblivious heuristics to prune the search space and make
    validation possible in practice.
  \item We implement the validation framework and evaluate it using the \ns{}
    network simulator~\cite{riley2010ns}. Our framework accurately identifies
    both synthetic and previously unknown violations in \ns{}'s implementations
    of the 802.11 and ARF protocols.  We also applied our framework to a
    commercial product under development and was able to found three latent bugs
    that were not observed previously.
\end{enumerate}

This paper is an extension of our work that appeared in RV 2016~\cite{Shi2016}.
We added the proof for Lemma~\ref{lem:mutation}, \ref{lem:sub} and
Theorem~\ref{the:equivalent}. We included new results that show the searching
cost (Fig.~\ref{fig:cost}), and details of the ARF protocol violations
(Section~\ref{subsec:arf}). Further, we reported the application of our
framework to a commercial product under development in
Section~\ref{subsec:xbox}.

\makeatletter{}\section{Background and Motivating Example}
\label{sec:motivation}

We encountered the uncertainty problem while testing the protocol implementation
of a popular wireless game controller. A custom wireless protocol
was designed to meet the low latency and low power consumption goals. As is
common industry practice, the protocol specification was then handed over to
wireless chipset vendors for implementation. However, neither implementation
details nor trace collection capabilities are provided in the shipped firmware
due to intellectual property constraints and device resource limitation. Hence
using sniffers to validate the protocol implementation is the only option.

We initially developed a tool to validate certain protocol properties over the
sniffer trace, yet often found unacceptable amount of false alarms due to
the incompleteness of the sniffer traces, making the tool virtually useless. It
was clear that we needed to account for sniffer uncertainty.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.6\textwidth]{dot11_tx_ta.pdf}
  \caption{\textbf{Monitor State Machine for 802.11 Transmitter.}}
  \label{fig:dot11_tx_ta}
\end{figure}

To better understand the incompleteness of sniffer trace, consider the IEEE
802.11 (also known as \wifi{}) transmitter (DUT) state machine shown in
Fig.~\ref{fig:dot11_tx_ta}. After the DUT sends $\it{DATA}_i$---a data packet with
sequence number $i$ ($s_0\rightarrow s_1)$, it starts a timer and waits for the
acknowledgment packet---$Ack$. The DUT either receives $Ack$ within time $T_o$
($s_1\rightarrow s_0$), or it sends $\it{DATA}'_i$---retransmission of $\it{DATA}_i$
($s_1\rightarrow s_2$). Similarly, the DUT either receives the $Ack$ within $T_o$
($s_2\rightarrow s_0$) or aborts transmission and moves on to next
packet\footnote{To represent the state machine succinctly, our example assumes
that the DUT retries at most once.} ($s_2\rightarrow s_1$).

Given a complete log of DUT's packet transmission and reception events, it is
trivial to feed such a log into the state machine in Fig.~\ref{fig:dot11_tx_ta}
and validate the correctness of DUT's protocol implementation. However, due to
DUT limitations we have described earlier, this complete log is not
available. As a result, we seek to validate the DUT implementation using
sniffers.

There are two fundamental properties in wireless communication that bring uncertainty to
sniffer's observation: packet loss and physical diversity. The sniffer could
either miss packets sent from or to the DUT due to packet loss, or overhear packets
that are sent to but missed by the DUT due to physical diversity.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.6\textwidth]{false_pos.pdf}
  \caption{\textbf{Uncertainty of Sniffer Observations.} $Tr_{OTA}$ is
    the chronological sequence of packets sent by the DUT and the receiver.
    $Tr_{DUT}$ is DUT's internal events. $Tr_1$ and $Tr_2$ are two examples of
    many possible sniffer traces.}
  \label{fig:sniffer_in_middle}
\end{figure}

Consider a correct packet exchange sequence shown in
Fig.~\ref{fig:sniffer_in_middle}. The DUT first sends $\it{DATA}_0$.  Suppose the
receiver receives $\it{DATA}_0$ and sends the $Ack$ which the DUT does not receive.
Eventually the DUT's timer fires and it sends $\it{DATA}'_0$.  This time the
$\it{DATA}'_0$ reaches receiver and the DUT also receives the $Ack$.

Now consider two possible traces that could have been overheard by a sniffer shown
in Fig.~\ref{fig:sniffer_in_middle}.
In first sniffer trace $Tr_1$ where the sniffer
\textit{overhears} the first $Ack$ packet, a validation \textit{uncertainty}
arises when the sniffer sees the $\it{DATA}'_0$: was the previous $Ack$ missed by the
DUT or is there a bug in DUT which causes it to retransmit even after receiving
the $Ack$?

Similarly, consider the second possible sniffer trace $Tr_2$
where both the $\it{DATA}'_0$ and $Ack$ packets were missed by the sniffer. During
this period of time, it appears the DUT neither receives $Ack$ for $\it{DATA}_0$ nor
sends $\it{DATA}'_0$. Again, without any additional information it is impossible to
disambiguate between the sniffer missing certain packets and a bug in DUT's
retransmission logic.

Informally, the question we set out to answer in this paper is: given the
protocol monitor state machine and the sniffer's observation with inherent
uncertainty, how to accurately validate that the DUT behaves as specified?

\makeatletter{}\section{Prerequisites and Problem Statement}
\label{sec:background}

\subsection{Packet, Trace and Monitor State Machine}
\label{subsec:basic}

The alphabet of the monitor state machine is the finite set of all valid packets
defined by the protocol, denoted as $\mathbb{P}$.  A packet is a binary string
of a finite number of bits, encoding interesting protocol attributes such as
\texttt{src}, \texttt{dest}, \texttt{type}, \texttt{flags}, and physical layer
information, such as \texttt{channel}, \texttt{modulation}, etc. The input of
the state machine then corresponds to a time-ordered sequence of packets.

\begin{definition}[Packet Trace]
  \sloppy{
  A \textit{packet trace} is a finite sequence of $(timestamp, packet)$ tuple:
  $[(t_1, p_1), (t_2, p_2),\ldots,(t_n, p_n)]$ where $t_i \in \mathbb{Z}^+$ is
  the \textit{discrete} timestamp and $p_i$ is the packet observed at time
  $t_i$. The timestamps are strictly monotonically increasing: $t_i < t_{i+1}$
  for $1 \le i < n$.
}
\end{definition}
In addition to timestamp monotonicity, we also require that adjacent packets do
not overlap in time, $t_{i+1}-t_i > \text{\texttt{airtime}}(p_i)$ for $1 \le i <
n$, where \texttt{airtime()} calculates the time taken to transmit a packet.
The timestamp represents the observer's local clock ticks, and need not to be
synchronized among devices.

We use \textit{timed automata}~\cite{alur1994theory} to model the expected
behaviors of the DUT. A timed automata is a finite state machine with timing
constraints on the transitions: each transition can optionally start one or more
timers, which can later be used to assert certain events should be seen before
or after the time out event. We refer the readers to~\cite{alur1994theory} for
more details about timed automata.

\begin{definition}[Monitor]
  \sloppy{    A protocol monitor state machine $S$ is a 7-tuple
    $\{\Sigma, \mathbb{S}, \mathbb{X}, s_0, C, E, G\}$, where:
  }
  \begin{itemize}
    \item $\Sigma = \mathbb{P}$ is the finite input alphabet.

    \item $\mathbb{S}$ is a non-empty, finite set of states. $s_0 \in
      \mathbb{S}$ is the initial state.

    \item $\mathbb{X}$ is the set of boolean variables. We use $v = \{x
      \leftarrow true/false\ |\ x \in \mathbb{X}\}$ to denote an assignment of
      the variables. Let $\mathbb{V}$ be the set of such values $v$.

    \item $C$ is the set of clock variables. A \textit{clock variable} can be
      reset along any state transitions. At any instant, reading a clock
      variable returns the time elapsed since last time it was reset.

    \item $G$ is the set of guard conditions defined inductively by
      \begin{align*}
        g := true\ |\ c \le T\ |\ c \ge T\ |\ x\ |\ \neg g\ |\ g_1 \land g_2
      \end{align*}      where $c \in C$ is a clock variable, $T$ is a constant, and $x$ is a
      variable in $\mathbb{X}$.  A transition can choose not to use guard
      conditions by setting $g$ to be $true$.

    \item $E \subseteq \mathbb{S} \times \mathbb{V} \times \mathbb{S} \times
      \mathbb{V} \times \Sigma \times  G \times \mathscr{P}(C)$ gives the set of
      transitions.\\ $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$
      represents that if the monitor is in state $s_i$ with variable assignments
      $v_i$, given the input tuple $(t, p)$ such that the guard $g$ is
      satisfied, the monitor can transition to a state $s_j$ with variable assignments
      $v_j$, and reset the clocks in $C'$ to 0.
  \end{itemize}
  \label{def:sm}
\end{definition}

A tuple $(t_i, p_i)$ in the packet trace means the packet $p_i$ is presented to
the state machine at time $t_i$. The monitor {\it rejects a trace} $Tr$ if there
exists a prefix of $Tr$ such that all states reachable after consuming the
prefix have no valid transitions for the next $(t, p)$ input.

As an example, the monitor state machine illustrated in
Fig.~\ref{fig:dot11_tx_ta} can be formally defined as follows:

\begin{itemize}
  \item $\Sigma = \{\it{DATA}_i, \it{DATA}'_i, Ack\ |\ 0 \le i < N\}$.
  \item Clock variables $C = \{c\}$. The only clock variable $c$ is
    used for acknowledgment time out.
  \item $\mathbb{X} = \{i\}$, as a variable with ${\mathit log}(N) + 1$ bits to
    count from $0$ to $N$.  
  \item Guard constraints $G = \{ c \le T_o, c > T_o, T_o < c \le T_m\}$.
    $T_o$ is the acknowledgment time out value, and $T_m >
    T_o$ is the maximum delay allowed before the retransmission packet gets
    sent. $T_o$ can be arbitrary large but not infinity in order to check the
    liveness of the DUT.
\end{itemize}

The monitor state machine defines a \textit{timed language} $L$ which consists
of all valid packet traces that can be observed by the DUT.  We now give the
definition of protocol \textit{compliance} and \textit{violation}.

\begin{definition}[Violation and Compliance]
  Suppose $\mathbb{T}$ is the set of all possible packet traces collected from
  DUT, and $S$ is the state machine specified by the protocol. The DUT
  \textit{violates} the protocol specification if there exists an
  packet trace $Tr \in \mathbb{T}$ such that $S$ rejects $Tr$.
  Otherwise, the DUT is \textit{compliant} with the specification.
\end{definition}

The focus of this paper is to determine whether a \textit{given} sniffer trace
$Tr$ is evidence of a violation.

\subsection{Mutation Trace}
\label{subsec:mutation}

As shown in the motivation example in Fig.~\ref{fig:sniffer_in_middle}, a
sniffer trace may either miss packets that are present in DUT trace, or contain
extra packets that are missing in DUT trace. Note that in the latter case, those
extra packets must be all sent \textit{to} the DUT. This is because it is
impossible for the sniffer to overhear packets sent from the DUT that were not
actually sent by the DUT.

We formally capture this relationship with the definition of mutation trace.

\begin{definition}[Mutation Trace]
  \label{def:mutation}
  A packet trace $Tr'$ is a mutation of sniffer trace $Tr$ w.r.t a DUT if for
  all $(t, p) \in Tr \setminus Tr'$, $p.dest = DUT$, where $p.dest$ is the
  destination of packet $p$.
\end{definition}

By definition, either $Tr' \supseteq Tr$ (hence $Tr \setminus Tr' = \emptyset$),
or those extra packets in $Tr$ but not in $Tr'$ are all sent to the DUT. Note
that $Tr'$ may contain extra packets that are either sent to or received by
the DUT.

A mutation trace $Tr'$ represents a \textit{guess} of the corresponding DUT
packet trace given sniffer trace $Tr$.  In fact, the DUT packet trace must
be one of the mutation traces of the sniffer trace $Tr$.

\begin{lemma}
  Let $Tr_{DUT}$ and $Tr$ be the DUT and sniffer packet trace captured during
  the same protocol operation session, and $\mathcal{M}(Tr)$ be the set of
  mutation traces of $Tr$ with respect to DUT, then $Tr_{DUT} \in \mathcal{M}(Tr)$.
  \label{lem:mutation}
\end{lemma}\begin{proof}
  Let $\Delta = Tr \setminus Tr_{DUT}$ be the set of packets that are in $Tr$
  but not in $Tr_{DUT}$. Recall that it is not possible for the sniffer to
  observe packets sent from the DUT that the DUT did not send. Therefore,
  all packets in $\Delta$ are sent \textit{to} the DUT. That is, for
  all $(t, p) \in \Delta,\ p.dest = DUT$. By Definition~\ref{def:mutation},
  $Tr_{DUT} \in \mathcal{M}(Tr)$.
  \qed
\end{proof}

\subsection{Problem Statement}
\label{subsec:problem}

Lemma~\ref{lem:mutation} shows that $\mathcal{M}(Tr)$ is a \textit{complete} set
of guesses of the DUT packet trace. Therefore, the problem of validating DUT
implementation given a sniffer trace can be formally defined as follows:

\begin{problem}
  \label{prob:validation}
  VALIDATION\\
  \textbf{instance} A protocol monitor state machine $S$ and a sniffer trace $Tr$.\\
  \textbf{question} Does there exist a mutation trace $Tr'$ of $Tr$ that satisfies $S$?
\end{problem}

If the answer is no, a definite violation of the DUT implementation can be
claimed. Nevertheless, if the answer is yes, $S$ may still reject $Tr_{DUT}$.
In other words, the conclusion of the validation can either be
\textit{definitely wrong} or \textit{probably correct}, but not
\textit{definitely correct}.  This is the fundamental limitation caused by the
uncertainty of sniffer traces.

\makeatletter{}\section{Validation Framework}
\label{sec:framework}

\makeatletter{}\subsection{Augmented State Machine}
\label{subsec:augment}

To deal with the inherent uncertainty of sniffer traces, we propose to
systematically augment the original monitor state machine with non-deterministic
transitions to account for the difference between the sniffer and DUT traces.

Before formally defining the augmented state machine, we first use an example to
illustrate the basic idea. Fig.~\ref{fig:augment} shows the augmented state
machine for 802.11 transmitter state machine shown in
Fig.~\ref{fig:dot11_tx_ta}.  For each existing transition (e.g., $s_0\rightarrow
s_1$), we add an \textit{empty transition} with same clock guards and resetting
clocks.  This accounts for the possibility when such packet was observed by
the DUT but missed by the sniffer.  Additionally, for each transition triggered
by a \textit{receiving} packet (i.e., $p.dest = DUT$), such as $s_1\rightarrow
s_0$ and $s_2\rightarrow s_0$, we add a \textit{self transition} with the same
trigger packet and clock guards, but an empty set of resetting clocks and no
assignments to variables. This allows the state machine to make progress when
the sniffer missed such packets.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.7\textwidth]{dot11_tx_checker.pdf}
  \caption{\textbf{Augmented Monitor State Machine.} Augmented transitions are
  highlighted in bold face. $\overline{Pkt}$ means either $\epsilon$ or $Pkt$.}
  \label{fig:augment}
\end{figure}

There are two things to note. First, self transitions are added only for
packets sent \textit{to} the DUT, since the sniffer will not overhear packets
\textit{from} the DUT if they were not sent by the DUT. Second, no augmented
transitions are added for the packets that are sent to DUT yet are missed by both
the DUT and the sniffer, since such packets do not cause difference between the
DUT and sniffer traces.

The augmented state machine in Fig.~\ref{fig:augment} will accept the sniffer
packet traces $Tr_1$ and $Tr_2$ shown in Fig.~\ref{fig:sniffer_in_middle}.  For
instance, one accepting transition sequence on sniffer trace $Tr_1$ is
$s_0\rightarrow s_1 \rightarrow_s s_1\rightarrow s_2 \rightarrow s_0$, and the
sequence for $Tr_2$ is $s_0 \rightarrow s_1 \rightarrow_e s_2 \rightarrow s_0$,
where $\rightarrow$ is the transition from the original state machine,
$\rightarrow_e$ and $\rightarrow_s$ are the augmented empty and self transitions
respectively.

We now formally define the augmented state machine.

\begin{definition}[Augmented Monitor]
  \sloppy{
  An augmented state machine $S^+$ for a monitor state machine $S$ is a 7-tuple
  $\{\boldsymbol{\Sigma^+}, \mathbb{S}, \mathbb{X}, s_0, C, \boldsymbol{E^+},
  G\}$, where $\mathbb{S}, \mathbb{X}, s_0, C, G$ are the same as $S$.
  $\Sigma^+=\{\epsilon\} \cup \Sigma$ is the augmented input alphabet with the
  empty symbol, and $E^+ \supset E$ is the set of transitions, which includes:
}
  \begin{itemize}
    \item $E$: existing transitions (\textbf{Type-0}) in $S$.
    \item $E^+_1$: empty transitions (\textbf{Type-1}) for transitions in $E$.
    \item $E^+_2$: self transitions \textbf{(Type-2)} for transitions
      triggered by receiving packets.
  \end{itemize}
\end{definition}

\begin{figure}[t!]
\begin{algorithm}[H]
  \caption{Obtain Augmented Transitions $E^+$ from $E$}
  \label{alg:augment}
  \begin{algorithmic}[1]
    \Function{augment}{$E$}
    \Let{$E^+$}{$\emptyset$}
    \ForAll{$\langle s_i, v_i, s_j, v_j, p, g, C'\rangle \in E$}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, p, g, C'\rangle\}$\Comment{Type-0}}
      \label{alg:augment:type0}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, s_j, v_j, \boldsymbol{\epsilon}, g, C'\rangle\}$\Comment{Type-1}}
      \label{alg:augment:type1}
      \If{$p.dest = DUT$}
      \Let{$E^+$}{$E^+ \cup \{\langle s_i, v_i, \boldsymbol{s_i},
      \boldsymbol{v_i}, p, g, \boldsymbol{\emptyset}\rangle\}$\Comment{Type-2}}
        \label{alg:augment:type2}
      \EndIf
    \EndFor
    \State \Return{$E^+$}
    \EndFunction
  \end{algorithmic}
\end{algorithm}
\end{figure}

Algorithm~\ref{alg:augment} describes the process of transforming $E$ into
$E^+$. In particular, Line~\ref{alg:augment:type0} adds existing transitions in
$E$ to $E^+$, while line~\ref{alg:augment:type1} and~\ref{alg:augment:type2} add
Type-1 and Type-2 transitions to $E^+$ respectively.  We have highlighted the
elements of the tuple that differ from the underlying Type-0 transition. Note
that in Type-2 transitions, both the state and the variables stay the same after
the transition.

With augmented state machine $S^+$, we can use Type-1 transitions to
non-deterministically infer packets missed by the sniffer, and use Type-2
transitions to consume extra packets captured by the sniffer but missed by the
DUT.

A accepting run of $S^+$ on sniffer trace $Tr$ yields a mutation trace $Tr'$
which represents one possibility of the DUT trace. Specifically, $Tr'$ can be
obtained by adding missing packets indicated by Type-1 transitions to $Tr$, and
removing extra packets indicated by Type-2 transitions from $Tr$

We show that the VALIDATION problem is equivalent to the
satisfiability problem of $Tr$ on $S^+$.

\begin{theorem}
  There exists a mutation trace $Tr' \in \mathcal{M}(Tr)$ that satisfies $S$ if
  and only if $Tr$ satisfies $S^+$.
 \label{the:equivalent}
\end{theorem}
\begin{proof}
  Assume $Tr$ satisfies $S^+$, and $P$ is a sequence of accepting transitions,
  we construct a mutation trace $Tr'$ using $P$ and show that $Tr'$ satisfies
  $S$.

  \sloppy{    Initially, let $Tr'=Tr$, then for each \textit{augmented} transition $\langle s_i,
  v_i, s_j, v_j, \sigma, g, C'\rangle \in P$:
  }
  \begin{itemize}
    \item If this is a Type-1 transition, add $(t, p)$ to $Tr'$, where $t$ is a
      timestamp that satisfies $g$ and $p$ is the missing packet.
    \item If this is a Type-2 transition, remove corresponding $(t, p)$ from
      $Tr'$.
  \end{itemize}
  It is obvious that $Tr'$ is a mutation trace of $Tr$, since only receiving
  packets are removed in the process.

  Now we show that there exists a accepting transition sequence $P'$ of $S^+$ on
  input $Tr'$ that does not contain augmented transitions.  In particular, $P'$
  can be obtained by substituting all Type-1 transitions with corresponding
  original transitions, and removing all Type-2 transition.  Since $P'$ does not
  contain augmented transitions, it is also an accepting transition sequence of
  $S$ on input $Tr'$, hence $Tr'$ satisfies $S$.

  On the other hand, assume $Tr' \in \mathcal{M}(Tr)$ and $Tr'$ satisfies $S$.
  Suppose $P'$ is the accepting transition sequences of $S$ on input $Tr'$.
  We first note that $P'$ is also the accepting transitions of $S^+$ on input
  $Tr'$, since $E \subset E^+$.

  We construct a accepting transition sequence $P$ of $S^+$ on input $Tr$ as
  follows.
  \begin{itemize}
    \item For each packet $p \in Tr' \setminus Tr$, substitute the transition
      $\langle s_i, v_i, s_j, v_j, p, g, C'\rangle$ with the corresponding Type-1
      transition $\langle s_i, v_i, s_j, v_j, \epsilon, g, C'\rangle$.
    \item For each transition $\langle s_i, v_i, s_j, v_j, \sigma, g, C'\rangle$
      followed by packet $p \in Tr\setminus Tr'$, add a Type-2 self
      transition $\langle s_j, v_j, s_j, v_j, p, g, \emptyset\rangle$. This is
      possible since $Tr'$ is a mutation trace of $Tr$, thus  for all $p \in Tr'
      \setminus Tr$, $p.src \ne DUT$.
  \end{itemize}
  Therefore, $Tr$ satisfies $S^+$.
  \qed
\end{proof}

By Theorem~\ref{the:equivalent}, the inherent uncertainty of the sniffer traces
is explicitly represented by the augmented transitions, and can be
systematically explored using the well established theory of state machine.

\begin{comment}
One immediate observation can be drawn from Theorem~\ref{the:equivalent} by
contradiction.

\begin{corollary}
  If $S^+$ rejects $Tr$, then $S$ rejects $Tr_{DUT}$.
  \label{cor:false_pos}
\end{corollary}
In the context of validation where we raise a violation alarm when $S^+$
rejects $Tr$, Corollary~\ref{cor:false_pos} provides a {\it sufficient} condition for reporting incorrect traces. 
However, when $S^+$ accepts $Tr$, $S$ could still reject $Tr_{DUT}$.
In other words, the conclusion of the validation can either be \textit{definitely
wrong} or \textit{possibly correct}, but not \textit{definitely correct}.
This is the fundamental limitation caused by the uncertainty of sniffer traces.
\end{comment}

\makeatletter{}\subsection{Problem Hardness}
\label{subsec:hard}

In this section, we show that the VALIDATION problem is NP-complete. In
fact, the problem is still NP-complete even with only one type of augmented
transitions.

Recall that Type-1 transitions are added because the sniffer may miss packets.
Suppose an imaginary sniffer that is able to capture \textit{every} packet ever
transmitted, then only Type-2 transitions are needed since the sniffer may still
overhear packets sent to the DUT. Similarly, suppose another special sniffer
that would not overhear any packets sent to the DUT, then only Type-1
transitions are needed to infer missing packets.

We refer the augmented state machine that only has Type-0 and Type-1
transitions as $S^+_1$, and the augmented state machine that only has Type-0 and
Type-2 transitions as $S^+_2$. And we show that each subproblem of determining
trace satisfiability is NP-complete.

\begin{problem}
  VALIDATION-1\\
  Given that $Tr\setminus Tr_{DUT}=\emptyset$ (sniffer does not overhear
  packets).\\
  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
  \textbf{question} Does $S^+_1$ accept $Tr$?
\end{problem}

\begin{problem}
  VALIDATION-2\\
  Given that $Tr_{DUT} \subseteq Tr$ (sniffer does not missing packets).\\
  \textbf{instance} Checker state machine $S$ and sniffer trace $Tr$.\\
  \textbf{question} Does $S^+_2$ accept $Tr$?
\end{problem}

\begin{lemma}
  \label{lem:sub}
  Both VALIDATION-1 and VALIDATION-2 are NP-complete.
\end{lemma}
\begin{proof}
  First, note that the length of mutation trace $Tr'$ is polynomial to the
  length of $Tr$ because of the discrete time stamp and non-overlapping packets
  assumption.
  Therefore, given a state transition sequence as witness, it can be verified in
  polynomial time whether or not it is an accepting transition sequence, hence
  both VALIDATION-1 and VALIDATION-2 are in NP.

  Next, we show how the SAT problem can be reduced to either one of the two
  problems.
  Consider an instance of SAT problem of a propositional formula $F$ with $n$
  variables $x_0,x_1,\ldots, x_{n-1}$, we construct a corresponding protocol and
  its monitor state machine as follows.

  The protocol involves two devices: the DUT (transmitter) and the endpoint
  (receiver).
  The DUT shall send a series of packets, $pkt_0, pkt_1,\ldots, pkt_{n-1}$.
  For each $pkt_i$, if the DUT receives the
  acknowledgment packet $ack_i$ from the endpoint, it sets boolean variable
  $x_i$ to be \texttt{true}.
  Otherwise $x_i$ remains to be \texttt{false}.
  After $n$ rounds, the DUT evaluate the formula $F$ using the assignments and
  sends a special packet, $pkt_{true}$, if $F$ is \texttt{true}.
  One round of the protocol operation can be completed in polynomial time since
  any witness of $F$ can be evaluated in polynomial time.

  \begin{figure}[t!]
    \centering
    \includegraphics[width=0.7\textwidth]{sat_sm.pdf}
    \caption{\textbf{Monitor State Machine for SAT Problem.}}
    \label{fig:sat}
  \end{figure}

  The protocol monitor state machine $S$ is shown in Fig.~\ref{fig:sat}.
  Initially, all $x_i$ is set to \texttt{false}.
  At state $s_0$, the DUT shall transmit $pkt_i$ within a unit time, transit to
  $s_1$ and reset the clock along the transition.
  At state $s_1$, either the DUT receives the $ack_i$ packet and
  set $x_i$ to be \texttt{true} ($s_1 \rightarrow s_0$), or the DUT continues to
  transmit the next packet $pkt_{i+1}$.
  After $n$ rounds, the state machine is $s_0$ or $s_1$ depending on whether
  $ack_{n-1}$ is received by the DUT.
  In either case, the DUT shall evaluate $F$ and transmit $pkt_{true}$ if $F$ is
  \texttt{true}.

  \sloppy{    Consider a sniffer trace $Tr_1=\{(0, pkt_0), (2, pkt_1), (4, pkt_2),\ldots,
  (2(n-1), pkt_{n-1}), (2n, pkt_{true})\}$.  That is, the sniffer only captures
  all $pkt_i$ plus the final $pkt_{true}$, but none of $ack_i$.  It is easy to see
  that $F$ is satisfiable if $S_1^+$ accepts $Tr_1$.  In particular, a successful
  run of $S_1^+$ on $Tr_1$ would have to guess, for each $pkt_i$, whether the
  Type-1 empty transitions should be used to infer the missing $ack_i$ packet,
  such that $F$ is \textit{true} at the end.  Note that for $Tr_1$, no Type-2 self
  transitions can be used since all packets in $Tr_1$ are sent from the DUT.
  Therefore, the SAT problem of $F$ can be reduced to the VALIDATION-1 problem of
  $S^+_1$ on sniffer trace $Tr_1$.
  }

>>>>

\section{Related Work}
\label{sec:related}
